---
title: "Fine-mapping simulated data with Bayesian linear regression"
author: "Peter Sørensen"
date: "8 October 2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

This example uses **gbase**, **glma**, **gmap** and **gbayes**, with the familiar
qgg preparation and MCMC arguments. The simulated panel contains 5,000 people
and 50,000 markers. The phenotype is shared with the
[linear and mixed-model example](Glma_linear_and_mixed_models_simulated_data.html).

## Prepare the data

```{r prepare}
library(gbase)
library(glma)
library(gmap)
library(gbayes)

setwd("C:/Users/Projects/Examples")
url <- "https://github.com/psoerensen/qgdata/raw/main/simulated_human_data/"
base <- "https://psoerensen.github.io/gact/Document/glma-simulated-data/"
download.file(paste0(url, "human.bed"), "human.bed", mode="wb")
download.file(paste0(base, "human.bim"), "human.bim")
download.file(paste0(url, "human.fam"), "human.fam")

Glist <- gprep(study="Simulation", bedfiles="human.bed",
  bimfiles="human.bim", famfiles="human.fam")
rsids <- gfilter(Glist=Glist, excludeMAF=0.05, excludeMISS=0.05,
  excludeCGAT=TRUE, excludeINDEL=TRUE, excludeDUPS=TRUE,
  excludeHWE=1e-12, excludeMHC=FALSE)
Glist <- gprep(study="Simulation", bedfiles="human.bed", rsids=rsids)

phenotypes <- read.csv(paste0(base, "example-phenotypes.csv"),
  colClasses=c("character", "numeric"))
y <- setNames(phenotypes$phenotype, phenotypes$id)
causal <- readLines(paste0(base, "causal-markers.txt"))
sets <- createMarkerSets(Glist=Glist, rsids=causal, upstream=1000, downstream=1000)

scan <- glma(y=y, Glist=Glist, method="linear")
stat <- checkStat(Glist=Glist, stat=scan$associations)
```

The supplied BIM corrects 13 invalid genetic-map entries without changing the
genotypes, physical positions or alleles. `upstream` and `downstream` count
markers within each chromosome.

## Fit the regions

```{r regional}
fit <- gmap(Glist=Glist, stat=stat, sets=sets, method="bayesC",
  nit=1000, nburn=100, nrun=2)
head(fit$post)
head(fit$conv)
head(fit$stat)
fit$cs
```

The default dense route computes full regional LD directly from BED. Posterior
tables include hyperparameters, effects and PIPs; convergence tables contain
native R-hat, ESS and status diagnostics. Inspect these before interpreting
posterior results. Credible-set grouping uses marginal PIPs.

## Fit genome-wide marker effects

```{r genome}
Glist <- gprep(Glist, task="sparseld", msize=1000, ldfiles="human.ld", nthreads=4)
fit <- gbayes(Glist=Glist, stat=stat, method="bayesC", nit=100, nrun=2)
head(fit$stat)
```

Genome-wide fitting reuses sparse LD. An eigen route is also available for
verified independent LD blocks; regional eigen fitting can be selected with
`gmap(..., algorithm="mcmc-eigen", eigen_threshold=0.99)` after sparse-LD
preparation. These routes have different likelihood and variance conventions;
the common argument names do not imply identical qgg posterior draws.

See [recorded tutorial timings](Tutorial_timings_and_memory.html).
