---
title: "Compute sparse LD and LD scores for 1000G"
author: "Peter Sørensen"
date: "8 October 2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

Use Glist from the [1000G preparation example](Process_1000G.html).
The reference already selects markers available in the gact database.

```{r sparse-ld}
library(gbase)
library(gact)
library(data.table)

GAlist <- readRDS("C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

for (ancestry in c("eur", "eas", "sas")) {
  descriptor <- file.path(GAlist$dirs["marker"],
    paste0("Glist_1000G_", ancestry, "_filtered.rds"))
  Glist <- readRDS(descriptor)
  Glist <- gprep(Glist, task="sparseld", msize=1000,
    ldfiles=file.path(GAlist$dirs["marker"], paste0("g1000_", ancestry, ".ld")),
    nthreads=4)
  saveRDS(Glist, descriptor)

  markers <- data.frame(rsids=unlist(Glist$rsids, use.names=FALSE),
    chr=unlist(Glist$chr), pos=unlist(Glist$pos),
    ea=unlist(Glist$a1), nea=unlist(Glist$a2),
    eaf=unlist(Glist$af), maf=unlist(Glist$maf), map=unlist(Glist$map))
  markers$ldscores <- getLDscores(Glist)[markers$rsids]
  fwrite(markers, file.path(GAlist$dirs["marker"],
    paste0("markers_1000G_", ancestry, "_filtered.txt.gz")))
}
```

`msize=1000` retains a 1,000-marker distance window, with no base-pair window
or r² cutoff. LD scores are computed in the same pass. To store fewer
correlations, supply an explicit `r2` cutoff. `ldfiles` names the output prefix;
gbase writes its own sparse-resource files. Keep those files with Glist for
analyses that read LD.

For eigenvalues of supplied marker sets, use `gprep(Glist, task="eigen", sets=sets)`.
See [genetic correlation](Genetic_correlation_CAD_T2D_gcorr.html) for methods
that reuse stored LD scores and regional moments.
