---
title: "Compute sparse LD matrices for 1000G"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This tutorial focuses on computing sparse Linkage Disequilibrium (LD) matrices for 1000 Genomes Project (1000G) data across different ancestries and exploring this data:   


```{r all-code, include=TRUE, eval=FALSE}
# Load libraries
library(qgg)
library(gact)
library(data.table)

# Load GAlist
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

# Save Glist for use later
Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G.rds"))


################################################################################
# Compute sparse LD matrices used for computing LD matrices in downstream analyses
################################################################################
# This step may take some tome to complete. In genereal it is recommended that R
# is linked to openblas, MKL or Atlas (default MacOS) for fast computation

# Identify high quality genetic markers in 1000G used for computation of sparse LD matrices
rsids <- gfilter(Glist = Glist,
                 excludeMAF = 0.05,
                 excludeMISS = 0.05,
                 excludeCGAT = TRUE,
                 excludeINDEL = TRUE,
                 excludeDUPS = TRUE,
                 excludeHWE = 1e-12,
                 excludeMHC = FALSE)

Glist <- gprep(Glist, task = "sparseld", msize = 1000, rsids = rsids, overwrite = TRUE)
saveRDS(Glist, file=file.path(GAlist$dirs["marker"],"Glist_1000G.rds"))

chr <- 1:22
markers <- data.frame(rsids=unlist(Glist$rsids[chr]),
                  chr=unlist(Glist$chr[chr]),
                  pos=unlist(Glist$pos[chr]),
                  ea=unlist(Glist$a1[chr]),
                  nea=unlist(Glist$a2[chr]),
                  eaf=unlist(Glist$af[chr]),
                  maf=unlist(Glist$maf[chr]),
                  map=unlist(Glist$map[chr]),
                  ldscores=NA)
rownames(markers) <- markers$rsids
markers[names(unlist(Glist$ldscores[chr])),"ldscores"] <- unlist(Glist$ldscores[chr])
fwrite(markers, file=file.path(GAlist$dirs["marker"],"markers_1000G.txt.gz"))


################################################################################
# Explore 1000G data
################################################################################

 Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G.rds"))

 chr <- 6
 plot(Glist$ldscores[[chr]], pch=".", frame.plot=FALSE,
      xlab="position", ylab="LD scores", main=paste("Chr",chr))

 df <- data.frame(rsids=unlist(Glist$rsids[chr]),
                  chr=unlist(Glist$chr[chr]),
                  pos=unlist(Glist$pos[chr]),
                  ea=unlist(Glist$a1[chr]),
                  nea=unlist(Glist$a2[chr]),
                  eaf=unlist(Glist$af[chr]),
                  maf=unlist(Glist$maf[chr]),
                  map=unlist(Glist$map[chr]),
                  ldscores=NA)
 rownames(df) <- df$rsids
 df[names(unlist(Glist$ldscores[chr])),"ldscores"] <- unlist(Glist$ldscores[chr])
 df <- na.omit(df)

 layout(matrix(1:2,ncol=2))
 boxplot(df$ldscores~df$ea)
 boxplot(df$maf~df$ea)

 layout(matrix(1:4,ncol=2))
 plot(y=df$ldscores,x=df$maf,pch=".", frame.plot=FALSE,
      xlab="MAF", ylab="LD scores", main=paste("Chr",chr))
 plot(y=df$ldscores,x=df$pos, pch=".", frame.plot=FALSE,
      xlab="Genome position (kb)", ylab="LD scores", main=paste("Chr",chr))
 plot(y=df$ldscores,x=df$map, pch=".", frame.plot=FALSE,
      xlab="Genome position (cM)", ylab="LD scores", main=paste("Chr",chr))
 plot(y=df$pos,x=df$map, pch=".", frame.plot=FALSE,
      xlab="Genome position (cM)", ylab="Genome position (kb)", main=paste("Chr",chr))



################################################################################
# Prepare Glist for 1000G data for different ancestries
################################################################################

library(qgg)
library(gact)
library(data.table)

# Load GAlist
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

# Marker IDs in database
rsids <- GAlist$rsids

Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_eur_filtered.rds"))

Glist <- gprep(Glist, task = "sparseld", msize = 1000, rsids = unlist(Glist$rsids), overwrite = TRUE)
saveRDS(Glist, file=file.path(GAlist$dirs["marker"],"Glist_1000G_eur_filtered.rds"))
file.remove(Glist$ldfiles)

markers <- data.frame(rsids=unlist(Glist$rsids),
                      chr=unlist(Glist$chr),
                      pos=unlist(Glist$pos),
                      ea=unlist(Glist$a1),
                      nea=unlist(Glist$a2),
                      eaf=unlist(Glist$af),
                      maf=unlist(Glist$maf),
                      map=unlist(Glist$map),
                      ldscores=unlist(Glist$ldscores))
rownames(markers) <- markers$rsids
fwrite(markers, file=file.path(GAlist$dirs["marker"],"markers_1000G_eur_filtered.txt.gz"))



#####################################
# Process EAS
#####################################

Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_eas_filtered.rds"))

Glist <- gprep(Glist, task = "sparseld", msize = 1000, rsids = unlist(Glist$rsids), overwrite = TRUE)
saveRDS(Glist, file=file.path(GAlist$dirs["marker"],"Glist_1000G_eas_filtered.rds"))
file.remove(Glist$ldfiles)

markers <- data.frame(rsids=unlist(Glist$rsids),
                      chr=unlist(Glist$chr),
                      pos=unlist(Glist$pos),
                      ea=unlist(Glist$a1),
                      nea=unlist(Glist$a2),
                      eaf=unlist(Glist$af),
                      maf=unlist(Glist$maf),
                      map=unlist(Glist$map),
                      ldscores=unlist(Glist$ldscores))
rownames(markers) <- markers$rsids
fwrite(markers, file=file.path(GAlist$dirs["marker"],"markers_1000G_eas_filtered.txt.gz"))

#####################################
# Process SAS
#####################################

Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_sas_filtered.rds"))

Glist <- gprep(Glist, task = "sparseld", msize = 1000, rsids = unlist(Glist$rsids), overwrite = TRUE)
saveRDS(Glist, file=file.path(GAlist$dirs["marker"],"Glist_1000G_sas_filtered.rds"))
file.remove(Glist$ldfiles)











