---
title: "Download and install gact database"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This tutorial is designed to download and set up the gact database, which is focused on genomic associations for complex traits. The tutorial is structured into two main parts: downloading and installing the database, and then providing an overview of its contents:   

<br>

## Download, save and reopen

The R package and database are installed separately. The default database contains
annotations, markers, gene and pathway statistics, Bayesian results, ATC
classifications and drug interactions. Genotype panels and LD preparation are
separate workflow steps.

```{r, eval=FALSE}
library(gact)
dbdir <- file.path(path.expand("~"), "gact-data")
GAlist <- gact(version = "hsa.0.0.1", dbdir = dbdir, task = "download")
# Check the existing download log before saving the database descriptor.
GAlist$downloads[, c("what", "status", "message")]
if (is.null(GAlist$downloads) || any(GAlist$downloads$status != "success")) {
  stop("Some components failed; inspect the download log before proceeding.")
}
descriptor <- file.path(GAlist$dbdir, paste0("GAlist_", GAlist$version, ".rds"))
saveRDS(GAlist, descriptor)
# In a later session, load the saved descriptor without starting downloads.
GAlist <- readRDS(file.path(dbdir, "hsa.0.0.1", "GAlist_hsa.0.0.1.rds"))
```

Choose a new version directory for the initial download. Repeating `gact()` against
an existing version directory stops; it does not resume installation. With the
returned `GAlist`, `downloadDB(GAlist, what = "component")` can retry a selected
component, which downloads that component again. Save the updated descriptor
after a successful retry. Keep the database at its original location because
the descriptor contains absolute paths.

In the recorded original setup, the basic database occupied approximately **3.04
GiB**, including retained archives, and its descriptor was approximately **50 MiB**.
The five Zenodo components totalled approximately **2.57 GiB of transfers**, with
ATC and drug interactions downloaded separately. The original setup operation took
**544.2 seconds (9 minutes 4 seconds)** and reached a process peak working set of
**2.68 GiB** on Windows with an Intel Core i7-1365U, about 16 GiB RAM, and R 4.4.1.
These are single-run measurements of that database snapshot, not minimum hardware
requirements or guaranteed timings. Allow additional storage for the descriptor,
genotype panels, LD resources and analysis outputs. Network speed and source
updates can change download times and database size.

## Overview of database contents

```{r overview, echo=TRUE, eval=FALSE}
GAlist$studies
GAlist$dirs
list.dirs(GAlist$dirs, recursive = TRUE)
list.files(GAlist$dirs["marker"])
list.files(GAlist$dirs["gstat"])
list.files(GAlist$dirs["gsea"])
```
