---
title: "Download and install gact database"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This tutorial is designed to download and set up the gact database, which is focused on genomic associations for complex traits. The tutorial is structured into two main parts: downloading and installing the database, and then providing an overview of its contents:   

<br>

```{r install_packages, echo=TRUE, eval=FALSE}

# library(devtools)
# devtools::install_github("psoerensen/gact")
# devtools::install_github("psoerensen/qgg")


# Download and install gact database
library(gact)
GAlist <- gact(version="hsa.0.0.1", dbdir="C:/Users/gact", task="download")
saveRDS(GAlist, file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")


# Overview of database content

GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

## GWAS Summary Statistics
GAlist$studies

## Accessing the 'dirs' Slot
GAlist$dirs

## Listing Directories and Files
### All Directories Recursively
list.dirs(GAlist$dirs, recursive = TRUE)

### Files under 'marker' Directory
list.files(GAlist$dirs["marker"])

### Files under 'gstat' Directory
list.files(GAlist$dirs["gstat"])

### Files under 'gsea' Directory
list.files(GAlist$dirs["gsea"])
