---
title: "Finemapping of gene regions using single trait Bayesian Linear Regression models"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This script is focusing on fine-mapping using the Bayesian Linear Regression (BLR) model. It processes and analyzes GWAS (Genome-Wide Association Studies) summary statistics and employs fine-mapping on specific gene regions.

```{r all-code, include=TRUE, eval=FALSE}
# Load libraries
library(qgg)
library(gact)

# Load GAlist
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")


# Check studies in gact database
GAlist$studies

# List BLR results files allready in gact database
list.files(GAlist$dirs["gbayes"])

# Load Glist with information on 1000G
Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G.rds"))

# Select study
studyID <- "GWAS1"

# Extract GWAS summary statistics for GWAS1
stat <- getMarkerStatDB(GAlist=GAlist, studyID=studyID)


# Check and align summary statistics based on marker information in Glist
stat <- checkStat(Glist=Glist, stat=stat,
                  excludeMAF=0.05,
                  excludeMAFDIFF=0.05,
                  excludeINFO=0.8,
                  excludeCGAT=TRUE,
                  excludeINDEL=TRUE,
                  excludeDUPS=TRUE,
                  excludeMHC=FALSE,
                  excludeMISS=0.05,
                  excludeHWE=1e-12)

# Extract gene-marker sets
sets <- getMarkerSetsDB(GAlist = GAlist, feature = "Genesplus")

genes <- c("ENSG00000099204","ENSG00000118972", "ENSG00000145996","ENSG00000140718")

# Fine-mapping using BLR model on selected gene regions
fit <- gmap(Glist=Glist, stat=stat, sets=sets[genes],
            method="bayesC", pi=0.001, h2=0.1,
            vb=0.01, ssb_prior=0.01, nub=5,
            checkConvergence=TRUE, ntrial=3,
            pruneLD=TRUE, r2=0.95, checkLD=FALSE,
            nit=1000, nburn=500, verbose=TRUE,
            updateB=TRUE, updatePi=FALSE, updateG=TRUE, updateE=TRUE)

# Posterior estimates of hyper-parameters (ve,vg,vb,pi,pip) for every fine-mapped region
head(fit$post)

# Convergence statistics for every fine-mapped region
head(fit$conv)

# Posterior estimates of marker effects for every fine-mapped region
head(fit$stat)

# Save fit object for all genes in directory of choice (here gbayes folder in database)
saveRDS(fit, file=file.path(GAlist$dirs["gbayes"],"fit_blr_genes_GWAS1.rds"))

