---
title: "Fine-mapping GWAS summary statistics"
author: "Peter Sørensen"
date: "8 October 2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

This example uses gact summaries and a matching EUR reference prepared with
gbase. gmap retains the familiar `sets`, `method`, `nit` and `nburn` arguments.

```{r prepare}
library(gact)
library(gbase)
library(gmap)
library(gscore)

GAlist <- readRDS("C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")
Glist <- readRDS(file.path(GAlist$dirs["marker"], "Glist_1000G_eur_filtered.rds"))
stat <- getMarkerStat(GAlist=GAlist, studyID="GWAS1")
stat <- checkStat(Glist=Glist, stat=stat, excludeMAF=0.05,
  excludeMAFDIFF=0.05, excludeINFO=0.8, excludeCGAT=TRUE,
  excludeINDEL=TRUE, excludeDUPS=TRUE, excludeMHC=FALSE,
  excludeMISS=0.05, excludeHWE=1e-12)

# Prepare the same marker selection used in the summary statistics.
Glist <- gprep(study=Glist$study, bedfiles=Glist$bedfiles,
  bimfiles=Glist$bimfiles, famfiles=Glist$famfiles, rsids=stat$rsids)
sets <- getMarkerSets(GAlist=GAlist, feature="Genesplus")
genes <- c("ENSG00000099204", "ENSG00000118972", "ENSG00000145996", "ENSG00000140718")
```

## Fine-map selected gene regions

```{r genes}
fit <- gmap(Glist=Glist, stat=stat, sets=sets[genes], method="bayesR",
  nit=1000, nburn=100, nrun=2, updateE=FALSE)
head(fit$post)
head(fit$conv)
head(fit$stat)
fit$cs
```

The dense route reads regional LD on demand. For GWAS summaries from a
different sample than the reference panel, residual variance learning is
disabled with `updateE=FALSE`.

## Reuse sparse LD for an eigen fit

```{r eigen}
Glist <- gprep(Glist, task="sparseld", msize=1000,
  ldfiles=file.path(GAlist$dirs["marker"], "t2d-finemapping.ld"), nthreads=4)
fit <- gmap(Glist=Glist, stat=stat, sets=sets[genes], method="bayesR",
  algorithm="mcmc-eigen", eigen_threshold=0.99,
  nit=1000, nburn=100, nrun=2, updateE=FALSE)
fit$cs
```

Use one eigen threshold per fit. A smaller retained mass changes the fitted
likelihood; inspect its diagnostics along with the posterior and convergence tables.

## LD-score marker sets and scoring

```{r ld-sets}
ldscores <- getLDscores(Glist, chr="17")
sets <- createLDsets(ldscores=ldscores, msize=200, maxsize=2000, nsplit=200)
fit <- gmap(Glist=Glist, stat=stat, sets=sets, method="bayesC",
  nit=1000, nburn=100, nrun=2, updateE=FALSE)
pgs <- gscore(Glist=Glist, stat=fit$stat)
head(pgs$scores)
```

`createLDsets()` supplies candidate regions at low rolling LD scores. It does
not certify independence for genome-wide block-eigen models. Scoring applies
the posterior effects to the reference individuals; it does not validate risk
prediction in an independent cohort.
