---
title: "Fine-mapping with single trait Bayesian Linear Regression models on simulated data"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This tutorial focus on fine-mapping using the Bayesian Linear Regression (BLR) model. It processes and analyzes GWAS (Genome-Wide Association Studies) summary statistics and employs fine-mapping on specific gene or genomic regions. Here we present the use of Bayesian linear regression (BLR) models for fine mapping approach using summary data. Allowing joint estimation of variants’ effects while accounting for LD among variants, our BLR models using Gibbs sampling technique, perform regularization and conduct variable selection.  

<br>


For individual-level regression and chromosome-excluded mixed-model association on this panel, see the [glma analysis tutorial](Glma_linear_and_mixed_models_simulated_data.html), including the frozen phenotype, results, timings and memory measurements.

#### Prepare simulated data used in fine-mapping examples 
```{r, include=TRUE, eval=FALSE}
#library(qgg)
library(devtools)
devtools::install_github("psoerensen/qgg")

# Set working directory
setwd("C:/Users/Projects/Examples")

# Download simulated genotype data
url <- "https://github.com/psoerensen/qgdata/raw/main/simulated_human_data/human.bed"
download.file(url = url, mode = "wb", destfile = "human.bed")
url <- "https://github.com/psoerensen/qgdata/raw/main/simulated_human_data/human.bim"
download.file(url = url, destfile = "human.bim")
url <- "https://github.com/psoerensen/qgdata/raw/main/simulated_human_data/human.fam"
download.file(url = url, destfile = "human.fam")

# Prepare Glist with information on genotypes stored in PLINK files
Glist <- gprep(study = "Simulation", 
               bedfiles = "C:/Users/Projects/Examples/human.bed", 
               bimfiles = "C:/Users/Projects/Examples/human.bim",
               famfiles = "C:/Users/Projects/Examples/human.fam")

# Filter markers with predefined thresholds for MAF, missingness, etc.
rsids <- gfilter(Glist = Glist, excludeMAF = 0.05, excludeMISS = 0.05,
                 excludeCGAT = TRUE, excludeINDEL = TRUE, excludeDUPS = TRUE, excludeHWE = 1e-12,
                 excludeMHC = FALSE)

# Compute sparse LD using filtered markers (may take some time if R is not linked to openblas, atlas or MKL)
Glist <- gprep(Glist, task = "sparseld", msize = 1000, 
               rsids = rsids, 
               ldfiles = "C:/Users/Projects/Examples/human.ld",
               overwrite = TRUE)

# Simulate phenotype using filtered markers
sim <- gsim(Glist=Glist, rsids=rsids)

# Create marker sets surrounding each index SNPs
sets <- createMarkerSets(Glist=Glist, rsids=sim$causal, upstream=1000, downstream=1000)

# Perform single marker regression analyses to get summary statistics
stat <- glma(y=sim$y,Glist = Glist, rsids=rsids)
```

<br>

##### Region-wide fine-mapping using simulated data 
```{r, include=TRUE, eval=FALSE}

# Fine-mapping in selected genome regions using a Bayes C prior for the marker variance
fit <- gmap(Glist=Glist, stat=stat, sets=sets, method="bayesC", 
            algorithm="mcmc", output="full",verbose=TRUE)


# Posterior estimates of hyper-parameters (ve,vg,vb,pi,pip) for every fine-mapped region
head(fit$post)

# Convergence statistics for every fine-mapped region
head(fit$conv)

# Posterior estimates of marker effects for every fine-mapped region
head(fit$stat)

# Credible sets for every fine-mapped region
head(fit$cs)
```

<br>

##### Genome-wide fine-mapping using simulated data 
```{r, include=TRUE, eval=FALSE}

# Fine-mapping across all chromosomes using a Bayes C prior for the marker variance
fit <- gbayes(Glist=Glist, stat=stat, method="bayesC", 
            algorithm="mcmc", verbose=TRUE)

# Posterior estimates of hyper-parameters (ve,vg,vb,pi,pip) for chromosomes in Glist
head(fit$post)

# Convergence statistics for chromosomes in Glist
head(fit$conv)

# Posterior estimates of marker effects for chromosomes in Glist
head(fit$stat)

# Credible sets for every fine-mapped region defined by sets
fit$cs <- getCredibleSets(Glist=Glist, fit=fit, sets=sets)
head(fit$cs)

```

## The same marker sets with standalone gsuite R packages

The original qgg examples above remain the reference workflow. Development
installations of gbase, gmap and gbayes provide familiar calls on the same
statistics and marker sets. Use statistics from these selected BED people
when enabling regional residual-variance learning.

```{r standalone-finemapping, include=TRUE, eval=FALSE}
# Reuse stat and sets from the simulation above; no new phenotype is simulated.
# This tutorial's BIM has negative genetic-map entries. Preserve the original
# and make a separate validated metadata copy (physical coordinates unchanged).
source_bim <- "C:/Users/Projects/Examples/human.bim"
validated_bim <- "C:/Users/Projects/Examples/human-validated-map.bim"
if (file.exists(validated_bim)) stop("Choose a new validated BIM filename")
bim <- read.table(source_bim, stringsAsFactors=FALSE)
bim[!is.finite(bim[[3]]) | bim[[3]] < 0, 3] <- 0
write.table(bim, validated_bim, quote=FALSE, row.names=FALSE, col.names=FALSE)

Gstandalone <- gbase::gprep(
  bedfiles="C:/Users/Projects/Examples/human.bed", bimfiles=validated_bim,
  famfiles="C:/Users/Projects/Examples/human.fam", rsids=stat$rsids)
Gstandalone$assembly <- "simulated-human-tutorial"

regional <- gmap::gmap(Glist=Gstandalone, stat=stat, sets=sets,
  method="bayesC", algorithm="mcmc", nit=1000, nburn=100,
  seed=10, output="full", verbose=TRUE)
head(regional$stat) # bm: posterior mean, dm: PIP
head(regional$post) # regional ve, derived vg, vb, pi, summed pip
head(regional$conv) # native R-hat/ESS/MCSE and diagnostic statuses
regional$cs        # named marginal-PIP groups
regional$timings   # per-region LD, preparation and sampling times

# Genome-wide regression reuses stored sparse LD. Its preparation is separate
# from fitting; use the same 1000-marker window and zero cutoff as above.
Gstandalone <- gbase::gprep(Gstandalone, task="sparseld",
  out_prefix="C:/Users/Projects/Examples/human-standalone-ld",
  max_distance_bp=0, max_distance_variants=1000, r2=0, nthreads=1)
genome <- gbayes::gbayes(Glist=Gstandalone, stat=stat,
  method="bayesC", algorithm="mcmc", nit=100, nburn=0,
  seed=10, verbose=TRUE)
head(genome$stat)
genome$post
genome$conv
genome$cs <- gmap::getCredibleSets(Glist=Gstandalone, fit=genome, sets=sets)
genome$cs
```

Dense regional MCMC computes full LD in memory using R's configured BLAS and
requires positive-definite LD; it performs no truncation or PSD repair.
`algorithm="mcmc-eigen"` preserves the explicit sparse-LD eigen route.
`gbase::mapSets()` and `gbase::getLD()` expose marker membership and full
regional correlation. The standalone grouping helper reproduces qgg CS2/CS3
ordering and strict LD-squared cutoffs without depending on qgg at runtime.
Marginal-PIP groups are not component credible sets with guaranteed coverage.

The familiar fields retain native meanings: genome-wide `bm` uses standardized
genotype/phenotype units, `post` has familiar summary rows and `conv` reports
native R-hat/ESS/MCSE rather than Geweke z scores. `post$vg` averages retained
joint-effect genetic variance and `post$pip` sums marker PIPs; no independent
genetic-variance hyperparameter draw is added. The `mcmc` reference-LD residual
adjustment defaults to .9. Both standalone BayesC/BayesR interfaces accept
`mcmc-eigen`; genome-wide fitting requires verified exhaustive Glist LD blocks
from `gbase::define_ld_blocks()` and uses retained-space residual normalization.
The 1000-marker sliding-window resource above must be given a valid block
partition before selecting that route. Single-chain fits are provisional. Reusing the same data and
iteration counts does not make the two implementations identical.
