---
title: "Individual-level Bayesian regression with BayesC and BayesR"
author: "Peter Sørensen"
date: "9 October 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

## Introduction

Fit BayesC and BayesR to the simulated individual-level panel and inspect
posterior parameter estimates. We use all individuals; prediction and the
historical eight-fit comparison are in the
[advanced example](Gbayes_individual_level_simulated_data_advanced.html).
Use **gbayes 0.1.5 or later** and a compatible gbase installation.
This page shows a runnable recipe; rendering does not fit the models.

## Prepare the data

Set the working directory to an existing folder of your choice. Download the
same panel, corrected marker map and phenotype used in the
[linear and mixed models example](Glma_linear_and_mixed_models_simulated_data.html).

```{r prepare}
library(gbase)
library(gbayes)

setwd("C:/Users/Projects/Examples")
url <- "https://github.com/psoerensen/qgdata/raw/main/simulated_human_data/"
base <- "https://psoerensen.github.io/gact/Document/glma-simulated-data/"
download.file(paste0(url, "human.bed"), "human.bed", mode="wb")
download.file(paste0(base, "human.bim"), "human.bim")
download.file(paste0(url, "human.fam"), "human.fam")

Glist <- gprep(study="Simulation", bedfiles="human.bed",
  bimfiles="human.bim", famfiles="human.fam")
rsids <- gfilter(Glist=Glist, excludeMAF=0.05, excludeMISS=0.05,
  excludeCGAT=TRUE, excludeINDEL=TRUE, excludeDUPS=TRUE,
  excludeHWE=1e-12, excludeMHC=FALSE)
Glist <- gprep(study="Simulation", bedfiles="human.bed", rsids=rsids)

phenotypes <- read.csv(paste0(base, "example-phenotypes.csv"),
  colClasses=c("character", "numeric"))
y <- setNames(phenotypes$phenotype, phenotypes$id)
data <- data.frame(id=Glist$ids, phenotype=as.numeric(y[Glist$ids]))
```

## Fit BayesC

```{r bayesc}
fit_c <- gbayes(data=data, Glist=Glist, trait="phenotype", method="bayesc")
fit_c
```

The defaults learn effect variance, residual variance and mixture weights.
They use two reproducible chains, 1000 burn-in and 2000 sampling sweeps,
thinning by five and full marker sweeps. Starting variances use the observed
phenotype variance; they do not use the known number of causal markers.
The phenotype is centred internally without rescaling, so estimates retain
its original units. These are starting settings, not a convergence guarantee.

## Inspect posterior summaries

```{r posterior}
fit_c$posterior
head(fit_c$estimates)
```

| Parameter | Meaning |
| --- | --- |
| VB (`effect_variance`) | Slab variance scale for standardized marker effects |
| VE (`residual_variance`) | Residual variance |
| VG (`genetic_variance`) | Realized genetic variance in this panel |
| VLE (`vle`) | Sum of individual-marker variance contributions |
| VLD (`vld`) | Sum of cross-marker contributions, VG − VLE |
| h² (`h2`) | Draw-wise VG / (VG + VE) |

The table reports posterior means, SDs, credible intervals, R-hat, bulk ESS
and Monte Carlo error. VLD can be negative. Posterior mean VG also differs
from the variance of scores calculated using posterior mean marker effects.

## Fit BayesR

```{r bayesr}
fit_r <- gbayes(data=data, Glist=Glist, trait="phenotype", method="bayesr")
fit_r
fit_r$posterior
```

BayesR uses several nonzero variance components. Its initial mixture weights
and its Dirichlet learning prior are separate settings. The fitted `$prior`
and `$control` show the resolved defaults; explicit settings remain available.
See the [defaults guide](https://psoerensen.github.io/gtools/gsuite/docs/gbayes-workflow.html#simple-individual-level-defaults)
for the initialization formulas and explicit-prior compatibility.

## Check diagnostics

```{r diagnostics}
fit_c$posterior[, c("parameter", "rhat", "ess_bulk", "mcse_mean")]
fit_r$posterior[, c("parameter", "rhat", "ess_bulk", "mcse_mean")]
```

Printing a fit flags estimated quantities with unavailable diagnostics,
R-hat above 1.01 or bulk ESS below 100. Inspect traces and continue sampling
when needed; passing this screen does not prove convergence. Check variance
and mixture quantities separately. The
[advanced investigation](Gbayes_individual_level_simulated_data_advanced.html#investigation-prediction-agreement-and-sampling-diagnostics)
shows why stable predictions can coexist with poor parameter mixing.

## Optional scheduling during burn-in

```{r scheduled}
fit_scheduled <- gbayes(data=data, Glist=Glist, trait="phenotype",
  method="bayesc",
  control=list(marker_schedule=list(enabled=TRUE, burnin_only=TRUE)))
fit_scheduled$posterior
```

Posterior sampling still visits every marker. Compare diagnostics before
interpreting differences between fits. The simulated independent-marker
panel illustrates the variance decomposition; it does not validate realistic
population LD.
