---
title: "Individual-level Bayesian regression with BayesC and BayesR"
author: "Peter Sørensen"
date: "9 October 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

## Introduction

This example fits BayesC and BayesR to individual-level genotypes and a
simulated phenotype. We inspect posterior variance summaries, marker effects
and predictions, then enable scheduled marker updates.

We use the same 5,000-person PLINK panel and four-causal-marker phenotype as
the [linear and mixed models example](Glma_linear_and_mixed_models_simulated_data.html).
The revised example uses **gbayes 0.1.4 or later**, with compatible gbase and gscore
installations and a native build containing the
[all-marker weight-update correction](https://github.com/psoerensen/gbayes/commit/c198232).
The code is a runnable recipe; rendering this page reads saved
results rather than fitting the models again.

## Prepare the data

Download the simulated panel, corrected marker map and phenotype. Existing
files are reused. The original genotypes are unchanged.

```{r prepare}
library(gbase)
library(gbayes)
library(gscore)

dir.create("gbayes-individual-example", showWarnings=FALSE)
setwd("gbayes-individual-example")
panel_url <- "https://github.com/psoerensen/qgdata/raw/main/simulated_human_data/"
example_url <- "https://psoerensen.github.io/gact/Document/glma-simulated-data/"
fetch <- function(url, file) {
  if (!file.exists(file)) download.file(url, file, mode="wb")
}
fetch(paste0(panel_url, "human.bed"), "human.bed")
fetch(paste0(example_url, "human.bim"), "human.bim")
fetch(paste0(panel_url, "human.fam"), "human.fam")
fetch(paste0(example_url, "example-phenotypes.csv"), "phenotypes.csv")

panel <- gprep(bedfiles="human.bed")
phenotypes <- read.csv("phenotypes.csv", colClasses=c("character", "numeric"))
y <- setNames(phenotypes$phenotype, phenotypes$id)

# Split people before marker QC or model fitting.
set.seed(20261009)
train_ids <- sample(panel$ids, 4000)
test_ids <- panel$ids[!panel$ids %in% train_ids]
train <- gprep(bedfiles="human.bed", ids=train_ids)
markers <- gfilter(train, maf=.05, missing=.05, hwe=1e-12,
  ambiguous=TRUE, indels=TRUE, mhc=FALSE)
train <- gprep(bedfiles="human.bed", ids=train_ids, rsids=markers)
test <- gprep(bedfiles="human.bed", ids=test_ids, rsids=markers)

# Standardize the phenotype using training samples only.
y_mean <- mean(y[train_ids])
y_sd <- sd(y[train_ids])
data <- data.frame(id=train$study_ids,
  response=as.numeric((y[train$study_ids]-y_mean)/y_sd))
m <- length(markers)
```

Training-only QC retained 37,994 markers in the recorded run. Both genotype
coding and phenotype scaling for prediction come from the training samples.

## Fit BayesC

BayesC uses a point mass at zero and one nonzero effect-variance component.
For this simulation, initialize the inclusion probability using its four
causal markers. Real analyses require prior choices without known truth.

```{r bayesc}
pi <- 4/m
vb <- .5/(m*pi)
prior_c <- list(inclusion_probability=pi,
  effect_variance=vb, residual_variance=.5,
  effect_variance_prior=list(df=4, scale=vb/2),
  residual_variance_prior=list(df=4, scale=.25), weight_prior=c(1, 1))
control <- list(burnin=1000, sampling_sweeps=2000, thinning=5,
  seeds=c(17, 29), threads=2, retain_diagnostics=TRUE,
  likelihood_dimension=length(train_ids)-1,
  estimate_effect_variance=TRUE, estimate_residual_variance=TRUE,
  estimate_weights=TRUE)

fit_c <- gbayes(data=data, Glist=train, trait="response",
  method="bayesc", prior=prior_c, control=control)
```

Because weight learning is enabled, `inclusion_probability` initializes the
chain; `weight_prior=c(1,1)` defines its diffuse Beta learning prior. The starting
four-marker count is not imposed on the posterior. BayesR's initial `weights`
and Dirichlet `weight_prior` are likewise separate choices.

## Inspect posterior summaries

The posterior table contains means, SDs and credible intervals, with R-hat,
effective sample size and Monte Carlo error for checking chain agreement.

```{r posterior}
columns <- c("parameter", "mean", "sd", "lower", "upper")
fit_c$posterior[, columns]

# Marker effect means, variances and posterior inclusion probabilities.
head(fit_c$estimates)
```

Check sampling diagnostics separately. The sweep counts above are a starting
point, not a convergence guarantee. Inspect the quantities needed for the
analysis; stable predictions do not imply stable mixture weights or variances.

```{r diagnostics}
fit_c$posterior[, c("parameter", "rhat", "ess_bulk", "mcse_mean")]

# Flag quantities needing further investigation (a screening rule).
subset(fit_c$posterior, estimated &
  (diagnostic_status != "available" | is.na(rhat) | is.na(ess_bulk) |
   rhat > 1.01 | ess_bulk < 100))
```

| Parameter | Meaning |
|---|---|
| `genetic_variance` (VG) | Total genetic variance of joint marker effects |
| `vle` | Diagonal marker contribution |
| `vld` | Cross-marker contribution; VG = VLE + VLD |
| `effect_variance` (VB) | Nonzero marker-effect variance scale |
| `residual_variance` (VE) | Residual variance parameter |
| `h2` | Heritability, VG / (VG + VE), calculated for each draw |

VLD can be negative. VB is a prior variance scale for marker effects and
differs from total genetic variance. The variance of predictions made using
posterior mean effects also differs from posterior mean VG.

These are the original exploratory BayesC variance summaries on the standardized
training phenotype scale. That recorded run used 200 burn-in and 500 sampling
sweeps, rather than the longer starting settings above:

```{r recorded-c, echo=FALSE, eval=TRUE, purl=FALSE}
saved <- read.csv("gbayes-individual-data/simple-posterior.csv")
knitr::kable(saved[saved$fit == "sparse_bayesc_full",
  c("parameter", "mean", "sd", "lower", "upper")], digits=4, row.names=FALSE)
```

VG, VLE, VLD and h² use every post-burn-in sweep from each original chain.
VB and VE use the retained parameter draws. Credible intervals use the
requested probability (95% by default); effect thinning does not thin the
BED variance rows. Continuation reports the new sampling batch.

## Fit BayesR

BayesR uses a zero component and several nonzero variance components. Here,
their multipliers are 0.01, 0.1 and 1. The component weights and variance
parameters are learned during sampling.

```{r bayesr}
weights <- c(1-pi, .7*pi, .25*pi, .05*pi)
gamma <- c(0, .01, .1, 1)
vb_r <- .5/(m*sum(weights*gamma))
prior_r <- list(weights=weights, variance_multipliers=gamma,
  effect_variance=vb_r, residual_variance=.5,
  effect_variance_prior=list(df=4, scale=vb_r/2),
  residual_variance_prior=list(df=4, scale=.25), weight_prior=rep(1, 4))

fit_r <- gbayes(data=data, Glist=train, trait="response",
  method="bayesr", prior=prior_r, control=control)
fit_r$posterior[, columns]
head(fit_r$estimates)
```

```{r recorded-r, echo=FALSE, eval=TRUE, purl=FALSE}
knitr::kable(saved[saved$fit == "sparse_bayesr_full",
  c("parameter", "mean", "sd", "lower", "upper")], digits=4, row.names=FALSE)
```

## Enable scheduled marker updates

Scheduled updates visit active or likely candidate markers frequently and
skip some low-probability null markers, with periodic full sweeps. Here we use
scheduling **only during burn-in**. Every posterior sampling sweep then visits
all markers and updates learned mixture weights. Compare diagnostics rather
than assuming that a longer run has converged.

```{r scheduled}
scheduled_control <- control
scheduled_control$marker_schedule <- list(enabled=TRUE, burnin_only=TRUE)

fit_c_scheduled <- gbayes(data=data, Glist=train, trait="response",
  method="bayesc", prior=prior_c, control=scheduled_control)
fit_r_scheduled <- gbayes(data=data, Glist=train, trait="response",
  method="bayesr", prior=prior_r, control=scheduled_control)
fit_c_scheduled$posterior[, columns]
fit_r_scheduled$posterior[, columns]
```

## Predict held-out individuals

Score using the allele frequencies and effects stored in the fitted weights,
then return predictions to the original phenotype scale.

```{r prediction}
scores <- gscore(fit_c$weights, test, scale_scores=FALSE)$scores[, 1]
predicted <- y_mean + y_sd*scores
observed <- y[names(scores)]
c(correlation=cor(observed, predicted),
  rmse=sqrt(mean((observed-predicted)^2)))
```

The original exploratory sparse-trait comparison below used scheduling during
both burn-in and sampling (`burnin_only=FALSE`). Its timing and visit reductions
therefore do not describe the safer burn-in-only recipe above:

```{r recorded-comparison, echo=FALSE, eval=TRUE, purl=FALSE}
comparison <- read.csv("gbayes-individual-data/fit-comparison.csv")
knitr::kable(comparison[startsWith(comparison$fit, "sparse_"),
  c("fit", "correlation", "rmse", "seconds", "fraction_marker_visits")],
  digits=3, row.names=FALSE)
```

The prediction results are encouraging. The two original sparse-fit chains'
posterior-mean predictions correlate above 0.999; their held-out accuracy is
also very similar. Some mixture weights nevertheless mix slowly. Additional
sampling improved sparse BayesR's variance diagnostics, while full polygenic
BayesR still showed poor mixing. Use prediction agreement, variance diagnostics
and mixture diagnostics as separate checks. This does not establish posterior
equivalence or guarantee convergence of individual marker effects.
Passing a variance-only screen does not qualify the joint posterior when
other parameters still show poor mixing.

For the 500-causal-marker trait, variance plots, known-effect reference,
full diagnostics and continuation, see the
[advanced eight-fit example](Gbayes_individual_level_simulated_data_advanced.html).
It also shows a polygenic BayesR case where scheduling visited every marker
and provided no speed gain. The independent simulated panel illustrates
empirical cross-marker contributions, not realistic population-LD behavior.
