---
title: "Genetic correlation with gcorr"
author: "Peter Sørensen"
date: "7 October 2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

<style>
table { width: 100%; font-variant-numeric: tabular-nums; }
th, td { padding: 9px 12px !important; }
thead { background: #f3f6f8; }
tbody tr:nth-child(even) { background: #f8fafb; }
</style>

## Introduction

This example estimates heritability and genetic correlation for type 2 diabetes
(GWAS1) and coronary artery disease (GWAS2) using **gcorr** and a gact database.
LDSC, GNOVA and SumHer reuse stored LD scores. HESS uses the BED reference to
prepare regional projections once, then reuses them for fitting.

## Prepare the data

Use the database and filtered EUR reference from the
[sparse-LD example](Compute_sparseLD_1000G.html).
The preparation helper aligns alleles, reuses the stored scores and keeps the
same marker selection and extracted `stat$n` as the website examples.

```{r prepare}
library(gact)
library(gbase)
library(gcorr)

GAlist <- readRDS("path/to/GAlist_hsa.0.0.1.rds")
Glist <- readRDS("path/to/Glist_1000G_eur_filtered.rds")

# Load the preparation helper and EUR region boundaries.
base <- "https://psoerensen.github.io/gact/Document/data/"
source(paste0(base, "gcorr_cad_t2d_example.R"))
regions_file <- tempfile(fileext=".bed")
download.file(paste0(base, "EUR-fourier-ac125e47.bed"), regions_file)
input <- prepare_gcorr_cad_t2d(GAlist, Glist, regions_file)
rm(GAlist, Glist)

stat <- input$statistics
ldscores <- input$ldscores
blocks <- input$blocks
```

## Genome-wide heritability and genetic correlation

```{r genome-wide}
# LD score regression.
fit_ldsc <- gcorr_ldsc(stat, ldscores, blocks,
  marker_count=input$reference_marker_count,
  marginal_intercept="free", maximum_chi_square=80)
fit_ldsc$estimates

# GNOVA, using the LDSC cross-trait intercept.
fit_gnova <- gcorr_gnova(stat, blocks, ldscores,
  overlap_intercepts=fit_ldsc$matrices$ldsc_intercept)
fit_gnova$estimates

# SumHer with an equal-marker baseline.
architecture <- matrix(unname(ldscores), ncol=1,
  dimnames=list(names(ldscores), "baseline"))
fit_sumher <- gcorr_sumher(stat, architecture,
  c(baseline=input$reference_marker_count), ldscores, blocks,
  reference_sample_size=input$Glist$n, inflation="multiplicative")
fit_sumher$estimates
```

Results are estimate ± standard error:

| Method | T2D h² | CAD h² | Genetic correlation |
| :--- | ---: | ---: | ---: |
| LDSC | 0.470 ± 0.025 | 0.311 ± 0.019 | 0.384 ± 0.028 |
| GNOVA | 0.498 ± 0.028 | 0.111 ± 0.019 | 0.604 ± 0.059 |
| SumHer baseline | 0.488 ± 0.029 | 0.362 ± 0.023 | 0.387 ± 0.027 |

The SumHer baseline is illustrative. GNOVA uses a unit marginal noise baseline;
LDSC and this SumHer fit estimate intercepts or scalings.

## Chromosomes as marker sets

Define each chromosome as a set of marker names. These disjoint sets can reuse
the ordinary LD scores, assuming negligible LD between chromosomes.

```{r chromosomes}
sets <- split(stat[[1]]$rsids, paste0("chr", stat[[1]]$chr))
counts <- lengths(sets)
scores <- gcorr_set_scores(stat[[1]]$rsids, ldscores, sets)
overlap <- diag(as.numeric(counts))
dimnames(overlap) <- list(names(sets), names(sets))

fit_sets <- gcorr_ldsc_partitioned(stat, scores, ldscores, counts, blocks,
  annotation_overlap=overlap, reference_sample_size=input$Glist$n,
  reference_marker_count=sum(counts),
  marginal_intercept="free", overlap_intercept=0)
head(fit_sets$estimates)
```

Here the cross-trait intercept is fixed to zero, assuming no sample overlap or
cross-trait confounding. Heritabilities and covariances sum across sets;
correlations do not. Other annotations need their corresponding tagging scores.

## Regional HESS and rho-HESS

```{r hess}
# Prepare all regions for both traits. No LD/eigenvector files are saved.
projection <- gcorr_prepare_hess(stat, input$Glist,
  regions=input$regions, nthreads=4)

# Fit heritability and genetic correlation from the prepared moments.
fit_hess <- gcorr_fit_hess(projection,
  shared_sample_size=0, phenotypic_correlation=0)
fit_hess$estimates[fit_hess$estimates$component == "total", ]

# Fit one trait using the same preparation.
fit_t2d <- gcorr_fit_hess(projection, "GWAS1")
```

For these data, raw CAD HESS heritability is poorly determined and the genetic
correlation exceeds one; that ratio is not interpretable as a biological
correlation. The estimates use the database's sample-size convention and are
not liability-scale heritabilities. Genomic-control adjustments remain
sensitivity analyses. See the [gcorr guide](https://psoerensen.github.io/gtools/gsuite/docs/summary-statistic-analysis.html)
for these options and assumptions.

## Time and memory

Recorded on an Intel Core i7-1365U with 16 GB RAM. Memory is the whole R process's
resident peak with the required analysis inputs loaded; data preparation,
loading and saving are outside the timed step. HESS preparation uses four threads.

| Step | Time | Peak resident memory |
| :--- | ---: | ---: |
| LDSC | 9.3 s | 3.02 GiB |
| GNOVA | 6.4 s | 3.09 GiB |
| SumHer baseline | 12.9 s | 3.53 GiB |
| Chromosome sets and fit | 29.9 s | 3.27 GiB |
| HESS preparation | 18 min 49 s | 3.18 GiB |
| HESS pair fit from prepared moments | 3.3–3.6 s | 0.43 GiB |
