---
title: " Gene analysis using VEGAS"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Gene analysis using VEGAS (Versatile Gene-based Association Study) is a computational method used in genomics to perform gene-based association tests by aggregating SNP (single nucleotide polymorphism) data within a gene. It was developed to address challenges in genome-wide association studies (GWAS), where individual SNP-level associations can sometimes lack power to detect genes related to complex traits or diseases. VEGAS increases the statistical power by combining information from multiple SNPs within a gene and tests the joint association of these SNPs with a phenotype. VEGAS is particularly useful in studies where individual SNP signals are too weak to reach genome-wide significance but may jointly point to gene-level associations indicative of genetic risk for complex traits or diseases.

This script is dedicated to gene analysis using the VEGAS (Versatile Gene-based Association Study) approach within the context of the gact and standalone gsuite packages in R. It is designed to perform gene-based association studies for multiple traits across different ancestries, specifically European (EUR), East Asian (EAS), and South Asian (SAS).

<br>

The database provides aligned marker statistics and gene annotations. `checkStat()`
applies the selected QC filters; `getMarkerSets()` selects available markers.
The existing Glist is reused, and VEGAS selects the required statistic rows internally.

#### Prepare for VEGAS gene analysis shown below
```{r, include=TRUE, eval=FALSE}
# Load libraries
library(gbase)
library(glma)
library(gact)

# Load GAlist with information on gact database
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

# Check studies in gact database
GAlist$studies

```


<br>

#### Perform VEGAS gene analysis for a single study of European ancestry (EUR)
```{r, include=TRUE, eval=FALSE}
# Select study
studyID <- "GWAS6"

# Load Glist with information on 1000G matched to the ancestry of GWAS data
Glist <- readRDS(file.path(GAlist$dirs["marker"],"Glist_1000G_eur_filtered.rds"))

# Get GWAS summary statistics from gact database
stat <- getMarkerStat(GAlist=GAlist, studyID=studyID)

# Check and align summary statistics based on marker information in Glist
stat <- checkStat(Glist=Glist, stat=stat)

# Get gene sets for the available checked markers.
sets <- getMarkerSets(GAlist=GAlist, feature="Genesplus", rsids=stat$rsids)

# Prepare regional eigenvalues and fit VEGAS.
Glist <- gprep(Glist, task="eigen", sets=sets, nthreads=4)
res <- vegas(Glist=Glist, stat=stat)
```

<br>

#### Perform VEGAS gene analysis for a single study of East Asian ancestry (EAS)
```{r, include=TRUE, eval=FALSE}
# Select study
studyID <- "GWAS7"

# Load Glist with information on 1000G matched to the ancestry of GWAS data
Glist <- readRDS(file.path(GAlist$dirs["marker"],"Glist_1000G_eas_filtered.rds"))

# Get GWAS summary statistics from gact database
stat <- getMarkerStat(GAlist=GAlist, studyID=studyID)

# Check and align summary statistics based on marker information in Glist
stat <- checkStat(Glist=Glist, stat=stat)

# Get gene sets for the available checked markers.
sets <- getMarkerSets(GAlist=GAlist, feature="Genesplus", rsids=stat$rsids)

# Prepare regional eigenvalues and fit VEGAS.
Glist <- gprep(Glist, task="eigen", sets=sets, nthreads=4)
res <- vegas(Glist=Glist, stat=stat)
```

<br>

#### Perform VEGAS gene analysis for multiple studies of European ancestry
```{r, include=TRUE, eval=FALSE}
# Select study
studyID <- c("GWAS1","GWAS2")

# Load Glist with information on 1000G (EUR)
Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_eur_filtered.rds"))

# Get GWAS summary statistics from gact database
stat <- setNames(lapply(studyID, function(id)
  checkStat(Glist=Glist, stat=getMarkerStat(GAlist=GAlist, studyID=id))), studyID)

# Get gene sets for markers available in both checked studies.
rsids <- Reduce(intersect, lapply(stat, `[[`, "rsids"))
sets <- getMarkerSets(GAlist=GAlist, feature="Genesplus", rsids=rsids)

# Prepare eigenvalues once for both traits and fit VEGAS.
Glist <- gprep(Glist, task="eigen", sets=sets, nthreads=4)
res <- vegas(Glist=Glist, stat=stat)
```
