---
title: "GSEA using BLR model derived gene-level statistics and MAGMA"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This script is designed for advanced gene set analysis, utilizing a variety of tools and methodologies. Its primary focus lies in the integration and assessment of gene-marker sets, along with a detailed exploration of gene-disease associations derived from text mining, knowledge bases, and experimental data. The analysis utilizes gene-marker relationships, gene-level statistics, and gene-disease associations to conduct comprehensive gene set evaluations. This allow an in-depth analysis helping interpreting and validating genomic associations.

```{r all-code, include=TRUE, eval=FALSE}
# Load libraries
library(qgg)
library(gact)
library(msigdbr)
library(data.table)


# Load GAlist with information on gact database
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")


# Select study
studyID <- "GWAS2"

# Load BLR result
stat <- fread(file=file.path(GAlist$dirs["gbayes"],paste0(studyID,"_stat_BayesC.txt.gz")), data.table=FALSE)
rownames(stat) <- stat$rsids

# Extract gene-marker sets
msets <- getMarkerSetsDB(GAlist = GAlist, feature = "Genesplus")
msets <- mapSetsDB(sets=msets,featureID=stat$rsids,index=TRUE)

# Compute gene-level statistics from BLR fit object
bm_gene_stat <- sapply(msets,function(x){sum(abs(stat$bm[x]))})
dm_gene_stat <- sapply(msets,function(x){sum(abs(stat$dm[x]))})
vm_gene_stat <- sapply(msets,function(x){sum(abs(stat$vm[x]))})

# Create data.frame and add annotation
df <- data.frame(bm=bm_gene_stat, dm=dm_gene_stat, vm=vm_gene_stat)
df <- addAnnotationDB(df=df)
head(df)

# Create data.frame, add annotation, hyperlinks and write to file
df <- data.frame(bm=bm_gene_stat, dm=dm_gene_stat, vm=vm_gene_stat)
df <- addAnnotationDB(df=df, hyperlinkEXCEL = TRUE)
head(df)
file_name <- file.path(GAlist$dirs["gbayes"],paste0(studyID,"_genes_BayesC.csv"))
write.csv2(df,file=file_name,row.names=FALSE)-
 
# Get gene sets for KEGG pathways from mSigDB
msigdb <- msigdbr(species = "human", category = "C2", subcategory = "CP:KEGG")
gsets <- split(msigdb$ensembl_gene, f=msigdb$gs_name)
gsets <- mapSets(gsets,names(bm_gene_stat), index=FALSE)

# Fit linear model (i.e. MAGMA procedure) using BLR gene-level statistics
resB <- qgg:::magma(stat=bm_gene_stat, sets=gsets)
resD <- qgg:::magma(stat=dm_gene_stat, sets=gsets)
resV <- qgg:::magma(stat=vm_gene_stat, sets=gsets)
head(resB)
head(resD)
head(resV)

################################################################################
# GSEA in cis-eQTL region derived from GTEx 
################################################################################
sets <- getMarkerSetsDB(GAlist = GAlist, feature = "GTExV8", threshold=0.01)
res <- gsea(stat, sets=sets)
str(res)

# Fit linear model (i.e. MAGMA procedure) using BLR marker-level statistics
dm <- stat$dm
names(dm) <- stat$rsids
resGTEx <- qgg:::magma(stat=dm, sets=sets)
head(resGTEx)


################################################################################
# GSEA in regulatory genomic features 
################################################################################
sets <- getMarkerSetsDB(GAlist = GAlist, feature = "Regulatory Categories")
res <- gsea(stat, sets=sets)
str(res)

# Fit linear model (i.e. MAGMA procedure) using BLR marker-level statistics
dm <- stat$dm
names(dm) <- stat$rsids
resREG <- qgg:::magma(stat=dm, sets=sets)
head(resREG)

# Compute regulatory-level statistics from BLR fit object
regsets <- getMarkerSetsDB(GAlist = GAlist, feature = "Regulatory Regions")
regsets <- mapSetsDB(sets=regsets,featureID=stat$rsids,index=TRUE)
bm_reg_stat <- sapply(regsets,function(x){sum(abs(stat$bm[x]))})
dm_reg_stat <- sapply(regsets,function(x){sum(abs(stat$dm[x]))})
vm_reg_stat <- sapply(regsets,function(x){sum(abs(stat$vm[x]))})

# Top ranking regulatory regions
head(sort(dm_reg_stat, decreasing=TRUE))

# Create data.frame, add annotation, and write to file
df <- data.frame(bm=bm_reg_stat, dm=dm_reg_stat, vm=vm_reg_stat)
df <- addAnnotationDB(df=df, feature="Regulatory")
head(df)
file_name <- file.path(GAlist$dirs["gbayes"],paste0(studyID,"_regulatory_BayesC.csv"))
write.csv2(df,file=file_name,row.names=FALSE)

################################################################################ 
# GSEA based on gene-diseases association in disease database for each KEGG pathway
################################################################################

# Get gene sets for feature from mSigDB (KEGG pathways)
msigdb <- msigdbr(species = "human", category = "C2", subcategory = "CP:KEGG")
sets <- split(msigdb$ensembl_gene, f=msigdb$gs_name)

# Enrichment test based on integrated information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "DiseaseGenes", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 50)

# Enrichment test based on textmining information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "DiseaseGenesTMplus", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 50)

# Enrichment test based on textmining information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "DiseaseGenesKBplus", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 50)

# Enrichment test based on experiment information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "DiseaseGenesEXPplus", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 50)


################################################################################
# GSEA based on gene-diseases association in GWAS catalog for each KEGG pathway
################################################################################

# Get gene sets for feature from mSigDB (KEGG pathways)
msigdb <- msigdbr(species = "human", category = "C2", subcategory = "CP:KEGG")
sets <- split(msigdb$ensembl_gene, f=msigdb$gs_name)

# Enrichment test based on integrated information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "GWAScatalog", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 10)

