---
title: "Genetic correlation with gcorr"
author: "Peter Sørensen"
date: "7 October 2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

<style>
table { width: 100%; font-variant-numeric: tabular-nums; }
th, td { padding: 9px 12px !important; }
thead { background: #f3f6f8; }
tbody tr:nth-child(even) { background: #f8fafb; }
</style>

## Introduction

This example estimates heritability and genetic correlation for type 2 diabetes
(GWAS1) and coronary artery disease (GWAS2) using **gcorr** and a gact database.
LDSC, GNOVA and SumHer reuse stored LD scores. HESS uses the BED reference to
prepare regional projections once, then reuses them for fitting.

## Prepare the data

Use the database and filtered EUR reference from the
[sparse-LD example](Compute_sparseLD_1000G.html). gact supplies the harmonized
statistics and sample sizes; Glist already contains the LD scores.
The short interfaces below require gcorr 0.1.7 or later.

```{r prepare}
library(gact)
library(gcorr)

GAlist <- readRDS("path/to/GAlist_hsa.0.0.1.rds")
Glist <- readRDS("path/to/Glist_1000G_eur_filtered.rds")

studyIDs <- c("GWAS1", "GWAS2", "GWAS6")
stat <- getMarkerStat(GAlist=GAlist, studyID=studyIDs)

# Keep the common three-study marker selection; fit T2D and CAD.
stat <- list(z=stat$z[, 1:2, drop=FALSE],
             n=stat$n[, 1:2, drop=FALSE])
```

## Genome-wide heritability and genetic correlation

```{r genome-wide}
fit_ldsc <- gcorr(Glist, stat, method="ldsc")
fit_ldsc$estimates

fit_gnova <- gcorr(Glist, stat, method="gnova")
fit_gnova$estimates

fit_sumher <- gcorr(Glist, stat, method="sumher")
fit_sumher$estimates
```

The wrapper matches marker names, reuses stored scores and supplies delete
blocks. LDSC estimates intercepts and excludes squared Z above 80. GNOVA uses
the LDSC cross-trait intercept by default; SumHer uses an equal-marker baseline
with multiplicative scaling. The extracted `stat$n` is unchanged.

For LDSC, scores can also be retrieved directly from gact:

```{r direct-ldsc}
ldscores <- getLDscoresDB(GAlist=GAlist, ancestry="EUR", version="1000G")
fit <- gcorr::ldsc(z=stat$z, n=stat$n, ldscores=ldscores, what="rg")
fit$rg
```

The direct interface leaves squared-Z filtering optional. Recorded example
results, with the cutoff of 80, are estimate ± standard error:

| Method | T2D h² | CAD h² | Genetic correlation |
| :--- | ---: | ---: | ---: |
| LDSC | 0.470 ± 0.025 | 0.311 ± 0.019 | 0.384 ± 0.028 |
| GNOVA | 0.498 ± 0.028 | 0.111 ± 0.019 | 0.604 ± 0.059 |
| SumHer baseline | 0.488 ± 0.029 | 0.362 ± 0.023 | 0.387 ± 0.027 |

The SumHer baseline is illustrative. GNOVA uses a unit marginal noise baseline;
LDSC and this SumHer fit estimate intercepts or scalings.

## Chromosomes as marker sets

Define each chromosome as a set of marker names. These disjoint sets can reuse
the ordinary LD scores, assuming negligible LD between chromosomes.

```{r chromosomes}
sets <- getMarkerSets(GAlist=GAlist, feature="Chromosomes")
fit_sets <- gcorr(Glist, stat, sets=sets, method="ldsc",
  set_model="grouped", overlap_intercept=0)
head(fit_sets$estimates)
```

Here the cross-trait intercept is fixed to zero, assuming no sample overlap or
cross-trait confounding. Heritabilities and covariances sum across sets;
correlations do not. Other annotations need their corresponding tagging scores.

## Regional HESS and rho-HESS

HESS additionally needs the BED reference and a named list of independent
LD-region marker sets (`regions`). See the [regional preparation guide](https://psoerensen.github.io/gtools/gsuite/docs/summary-statistic-analysis.html#regional-methods).
Retrieve allele-labelled frames for this step; HESS checks and aligns them to
the BED reference itself.

```{r hess}
library(gbase)

# Convert an older qgg descriptor using its existing BED files if needed.
if (!inherits(Glist, "gs_genotypes"))
  Glist <- gprep(bedfiles=Glist$bedfiles, bimfiles=Glist$bimfiles,
    famfiles=Glist$famfiles)

ids <- intersect(rownames(stat$z), rownames(stat$n))
ids <- intersect(ids, unlist(Glist$rsids, use.names=FALSE))
stat_hess <- setNames(lapply(c("GWAS1", "GWAS2"), function(id)
  getMarkerStat(GAlist=GAlist, studyID=id, rsids=ids)), c("GWAS1", "GWAS2"))

# regions contains the marker names for each independent LD region.
regions <- readRDS("path/to/EUR_LD_region_marker_sets.rds")
projection <- gcorr_prepare_hess(stat_hess, Glist, regions=regions, nthreads=4)
fit_hess <- gcorr_fit_hess(projection,
  shared_sample_size=0, phenotypic_correlation=0)
fit_hess$estimates[fit_hess$estimates$component == "total", ]

# Reuse the same moments for one trait, without recomputing eigenvectors.
fit_t2d <- gcorr_fit_hess(projection, "GWAS1")
```

For these data, raw CAD HESS heritability is poorly determined and the genetic
correlation exceeds one; that ratio is not interpretable as a biological
correlation. The estimates use the database's sample-size convention and are
not liability-scale heritabilities. Genomic-control adjustments remain
sensitivity analyses. See the [gcorr guide](https://psoerensen.github.io/gtools/gsuite/docs/summary-statistic-analysis.html)
for these options and assumptions.

## Time and memory

Recorded on an Intel Core i7-1365U with 16 GB RAM. Memory is the whole R process's
resident peak with the required analysis inputs loaded; data preparation,
loading and saving are outside the timed step. These timings measure the numerical
fits; a GNOVA workflow call also estimates its LDSC intercept unless one is
supplied. HESS preparation uses four threads.

| Step | Time | Peak resident memory |
| :--- | ---: | ---: |
| LDSC | 9.3 s | 3.02 GiB |
| GNOVA | 6.4 s | 3.09 GiB |
| SumHer baseline | 12.9 s | 3.53 GiB |
| Chromosome sets and fit | 29.9 s | 3.27 GiB |
| HESS preparation | 18 min 49 s | 3.18 GiB |
| HESS pair fit from prepared moments | 3.3–3.6 s | 0.43 GiB |
