---
title: "Gene ranking with ridge and Bayesian PoPS"
author: "Peter Sørensen"
date: "8 October 2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

PoPS predicts gene scores from pathway features. Here **gsea** fits each
model after excluding the chromosome of the genes it predicts. Features,
optional marginal selection and posterior fitting use the training chromosomes.

## Prepare gene scores and sets

```{r prepare}
library(gact)
library(gsea)

GAlist <- readRDS("C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")
scores <- getVEGAS(GAlist=GAlist, studyID="GWAS1")
annotation <- readRDS(file.path(GAlist$dirs["gsets"], "genesplus_annotation.rds"))
stat <- data.frame(gene=rownames(scores),
  chromosome=annotation$chr[match(rownames(scores), annotation$EnsemblID)],
  score=scores[,1])
sets <- getFeatureSets(GAlist=GAlist, feature="Pathways", minsets=50)
```

## Fit the models

```{r fit}
fit <- pops(stat=stat, sets=sets, method="rr", threshold=0.05)
head(fit$predictions)

fit <- pops(stat=stat, sets=sets, method="bayesC",
  pi=0.001, nit=5000, nburn=1000, nrun=2)
head(fit$predictions)
fit$folds

fit <- pops(stat=stat, sets=sets, method="bayesR",
  pi=0.001, nit=5000, nburn=1000, nrun=2)
head(fit$predictions)
head(fit$features)
```

Bayesian PoPS reuses the native gsea/gbayes sampler. BayesC has a point mass
and one nonzero variance; BayesR uses three nonzero variance scales. The
defaults learn effect variance, residual variance and mixture weights. Inspect
each chromosome's posterior and chain diagnostics in `fit$fits`.

These calls use independent genes as a working covariance. Supply
`sampling=list(gene_covariance=blocks)` for chromosome-contained covariance
blocks. Bayesian `threshold` is optional; when omitted all features that vary
in the training fold are fitted. Scores use training-centered features and
posterior mean coefficients. They are gene priorities, not causal probabilities
or individual polygenic scores. The chromosome-held-out prediction rule differs
from qgg's ordinary full-data `pops()` fit.
