---
title: "Polygenic Priority Score"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Here we introduce Polygenic Priority Score (PoPS) which is a method for prioritizing genes based on polygenic signals from genome-wide association studies (GWAS). By integrating GWAS summary statistics with functional genomic data, PoPS identifies genes most likely involved in complex traits or diseases. This approach leverages the polygenic nature of traits, considering many genetic variants across the genome, and scores genes based on their likelihood of contributing to the trait. PoPS is valuable for focusing research on genes with potential biological relevance, facilitating more targeted follow-up studies in complex trait genomics.

<br>

#### Prepare input data for Polygenic Priority Scoring 
```{r, include=TRUE, eval=FALSE}
# Load libraries
library(qgg)
library(gact)

# Load GAlist with information on gact database
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

# Check studies in gact database
GAlist$studies

# Select GWAS study IDs
studyIDs <- "GWAS1"

# Get VEGAS gene-level statistics (i.e. z-statistics)
stat <- getVEGAS(GAlist=GAlist, studyID=studyIDs)

# Get gene sets for Reactome pathways
sets <- getFeatureSets(GAlist = GAlist, feature = "Pathways", minsets=50)
```

<br>

#### Polygenic Priority Scoring using different procedures
```{r, include=TRUE, eval=FALSE}
# POPS based on linear model, thresholding and ridge regression (Original PoPS)
scoreRR <- pops(stat=stat, sets=sets, method="rr",
              threshold=0.05)
head(scoreRR)

# POPS based on linear model and BLR using BayesC prior
scoreC <- pops(stat=stat, sets=sets, method="bayesC",
             pi=0.001, nit=5000, nburn=1000)
head(scoreC)

# POPS based on linear model and BLR using BayesR prior
scoreR <- pops(stat=stat, sets=sets, method="bayesR",
             pi=0.001, nit=5000, nburn=1000)
head(scoreR)
```
