---
title: " Gene analysis using VEGAS"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This script is dedicated to gene analysis using the VEGAS (Versatile Gene-based Association Study) approach within the context of the gact and qgg frameworks in R. It is designed to perform gene-based association studies for multiple traits across different ancestries, specifically European (EUR), East Asian (EAS), and South Asian (SAS).

Script Workflow:

Library Installation and Setup:

The script starts with comments for installing gact and qgg packages, followed by loading these libraries.
It loads the GAlist object, which contains structured information about the genomic database.
Data Preparation and Marker Set Extraction:

It displays studies within the gact database and extracts gene-marker sets using getMarkerSetsDB, focusing on a feature called "Genesplus".
VEGAS Analysis for Multiple European Traits:

The script selects specific European studies and loads corresponding 1000 Genomes Project data for European ancestry.
High-quality genetic markers are identified using gfilter.
GWAS summary statistics are obtained for these studies and used in a VEGAS analysis.
The results are saved for future reference.


VEGAS Analysis: Central to the script, VEGAS analysis is used to investigate gene-level associations using GWAS summary statistics.
Data Filtering and Alignment: Critical steps involve filtering for quality markers and aligning GWAS data with the 1000 Genomes Project data.
Ancestry-Specific Analysis: The script demonstrates the flexibility to analyze data across different ancestries, adjusting the input data and parameters accordingly.
This script exemplifies a comprehensive approach to gene-based association analysis in genomic studies, utilizing sophisticated tools available in the gact and qgg packages to handle complex genetic data.


```{r all-code, include=TRUE, eval=FALSE}



library(qgg)
library(gact)


# Load GAlist with information on gact database
GAlist <- readRDS(file="/faststorage/project/ukbiobank/projects/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

# Check studies in gact database
GAlist$studies

# Extract gene-marker sets
markerSets <- getMarkerSetsDB(GAlist = GAlist, feature = "Genesplus")


################################################################################
# VEGAS for a multiple traits/studies (EUR)
################################################################################

# Select study
studyID <- paste0("GWAS",c(1:2, 4:6))

# Load Glist with information on 1000G (EUR)
Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_eur_filtered.rds"))

# Identify high quality genetic markers in 1000G
rsids <- gfilter(Glist = Glist,
                 excludeMAF = 0.01,
                 excludeMISS = 0.05,
                 excludeCGAT = TRUE,
                 excludeINDEL = TRUE,
                 excludeDUPS = TRUE,
                 excludeHWE = 1e-12,
                 excludeMHC = FALSE)

# Get GWAS summary statistics from gact database
p <- getMarkerStatDB(GAlist=GAlist, studyID=studyID,
                     what="p", rsids=rsids)
dim(p)

# Gene analysis using VEGAS
res <- qgg:::vegas(Glist=Glist, sets=markerSets, p=p, verbose=TRUE)
saveRDS(res, file=file.path(GAlist$dirs["gsea"],"vegas_eur_filtered.rds"))


################################################################################
# VEGAS for a single trait/study (EAS)
################################################################################

# Select study
studyID <- "GWAS7"

# Load Glist with information on 1000G (EAS)
Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_eas_filtered.rds"))

# Get GWAS summary statistics from gact database
stat <- getMarkerStatDB(GAlist=GAlist, studyID=studyID)

# Check and align summary statistics based on marker information in Glist
stat <- checkStat(Glist=Glist, stat=stat)

# Gene analysis using VEGAS
res <- qgg:::vegas(Glist=Glist, sets=markerSets, stat=stat, verbose=TRUE)
saveRDS(res, file=file.path(GAlist$dirs["gsea"],"vegas_eas_filtered.rds"))


################################################################################
# VEGAS for a single trait/study (SAS)
################################################################################

# Select study
studyID <- "GWAS8"

# Load Glist with information on 1000G (SAS)
Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_sas_filtered.rds"))

# Get GWAS summary statistics from gact database
stat <- getMarkerStatDB(GAlist=GAlist, studyID=studyID)
stat$ea <- toupper(stat$ea)
stat$nea <- toupper(stat$nea)

# Check and align summary statistics based on marker information in Glist
stat <- checkStat(Glist=Glist, stat=stat)

# Gene analysis using VEGAS
res <- qgg:::vegas(Glist=Glist, sets=markerSets, stat=stat, verbose=TRUE)
saveRDS(res, file=file.path(GAlist$dirs["gsea"],"vegas_sas_filtered.rds"))

