---
title: " Gene analysis using VEGAS"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This script is dedicated to gene analysis using the VEGAS (Versatile Gene-based Association Study) approach within the context of the gact and qgg frameworks in R. It is designed to perform gene-based association studies for multiple traits across different ancestries, specifically European (EUR), East Asian (EAS), and South Asian (SAS).

```{r all-code, include=TRUE, eval=FALSE}
# Load libraries
library(qgg)
library(gact)


# Load GAlist with information on gact database
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

# Check studies in gact database
GAlist$studies

# Extract gene-marker sets
markerSets <- getMarkerSetsDB(GAlist = GAlist, feature = "Genesplus")


################################################################################
# VEGAS for a multiple traits/studies (EUR)
################################################################################

# Select study
studyID <- paste0("GWAS",c(1:2, 4:6))

# Load Glist with information on 1000G (EUR)
Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_eur_filtered.rds"))

# Identify high quality genetic markers in 1000G
rsids <- gfilter(Glist = Glist,
                 excludeMAF = 0.01,
                 excludeMISS = 0.05,
                 excludeCGAT = TRUE,
                 excludeINDEL = TRUE,
                 excludeDUPS = TRUE,
                 excludeHWE = 1e-12,
                 excludeMHC = FALSE)

# Get GWAS summary statistics from gact database
p <- getMarkerStatDB(GAlist=GAlist, studyID=studyID,
                     what="p", rsids=rsids)
dim(p)

# Gene analysis using VEGAS
res <- qgg:::vegas(Glist=Glist, sets=markerSets, p=p, verbose=TRUE)
saveRDS(res, file=file.path(GAlist$dirs["gsea"],"vegas_eur_filtered.rds"))


################################################################################
# VEGAS for a single trait/study (EAS)
################################################################################

# Select study
studyID <- "GWAS7"

# Load Glist with information on 1000G (EAS)
Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_eas_filtered.rds"))

# Get GWAS summary statistics from gact database
stat <- getMarkerStatDB(GAlist=GAlist, studyID=studyID)

# Check and align summary statistics based on marker information in Glist
stat <- checkStat(Glist=Glist, stat=stat)

# Gene analysis using VEGAS
res <- qgg:::vegas(Glist=Glist, sets=markerSets, stat=stat, verbose=TRUE)
saveRDS(res, file=file.path(GAlist$dirs["gsea"],"vegas_eas_filtered.rds"))


################################################################################
# VEGAS for a single trait/study (SAS)
################################################################################

# Select study
studyID <- "GWAS8"

# Load Glist with information on 1000G (SAS)
Glist <- readRDS(file=file.path(GAlist$dirs["marker"],"Glist_1000G_sas_filtered.rds"))

# Get GWAS summary statistics from gact database
stat <- getMarkerStatDB(GAlist=GAlist, studyID=studyID)

# Check and align summary statistics based on marker information in Glist
stat <- checkStat(Glist=Glist, stat=stat)

# Gene analysis using VEGAS
res <- qgg:::vegas(Glist=Glist, sets=markerSets, stat=stat, verbose=TRUE)
saveRDS(res, file=file.path(GAlist$dirs["gsea"],"vegas_sas_filtered.rds"))

