---
title: "Summary-statistics imputation"
author: "Peter Sørensen"
date: "8 October 2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

## Prepare the data

This example checks and imputes Z-scores for type 2 diabetes (GWAS1) and
coronary artery disease (GWAS2). Both studies are processed in one call using
**gbase** and the aligned statistics in the gact database.
It requires gbase 0.1.3 or later and the updated gact package.

Load your saved GAlist and an EUR Glist prepared with `gprep()`, using the same
genome build and marker backbone. Its BED files must remain accessible.
For LD checking, use a matched reference with at least 5,000 individuals.
The paths below refer to your existing database and prepared reference.

```{r prepare}
library(gact)
library(gbase)

GAlist <- readRDS("path/to/GAlist_hsa.0.0.1.rds")
Glist <- readRDS("path/to/Glist_EUR_reference.rds")

studyIDs <- c("GWAS1", "GWAS2")
stat <- getMarkerStat(GAlist=GAlist, studyID=studyIDs, rm.na=FALSE)
```

`rm.na=FALSE` keeps markers with missing statistics. The eligible markers in
Glist define the output backbone; gact supplies study ancestry and marker identity.

## Check and impute

```{r impute}
stat <- checkStat(Glist, stat, ldcheck=TRUE, impute=TRUE, nthreads=4)
head(stat$z)
```

The function aligns alleles, applies marker QC, checks consistency with local
reference LD, and imputes missing or LD-rejected Z-scores. Accepted observations
are preserved. Predictions below the default quality threshold of `r2=0.6`
remain missing, and reference-QC exclusions are not restored.

LD is computed from BED in bounded windows. No stored sparse-LD matrices or
prepared eigenvalues are required, and no new LD files are written.

## Inspect the results

```{r diagnostics}
# Number of imputed scores for each study, and their reference prediction quality.
quality <- attr(stat, "imputation")
colSums(quality$imputed)
head(quality$r2)

# Original harmonized scores rejected by the LD consistency check.
head(attr(stat, "ldcheck")$original)
```

Known values in `stat$n` are retained; missing sample sizes are not guessed.
Effect estimates, standard errors and P values for new or repaired scores are
left missing. A later analysis must supply the inputs it needs and account for
imputation uncertainty. The LD check is a DENTIST-style diagnostic; its P values
are approximate and do not establish that the reference equals the original study LD.

## Other reference choices

For imputation without LD checking, use `ldcheck=FALSE` in the same call.
The filtered EUR Glist from the [1000G example](Compute_sparseLD_1000G.html)
can illustrate that route; its smaller sample size limits prediction quality.
For already aligned statistics, imputation alone is `stat <- impute(Glist, stat)`.
Start from freshly extracted statistics when comparing alternative settings.

A combined reference can be labelled with
`Glist <- setReferenceGroups(Glist, list(EUR=eur_ids, EAS=eas_ids))`, where
`eur_ids` and `eas_ids` are disjoint sample-ID vectors in that reference.
Gact's study ancestry then selects the corresponding group for each trait.
Explicit mixture weights are supported for independent standardized meta-analysis
scores with constant coefficients; they are not a general model of local admixture.
See the [preparation guide](https://psoerensen.github.io/gtools/gsuite/docs/genotype-preparation.html#summary-statistics-checking-and-imputation)
for ancestry selection, mixture weights and controls.
