---
title: "LD Score regression"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This script demonstrates how to estimate narrow-sense heritability and genetic correlation using the LD Score Regression (LDSC) method within the context of a gact database. The script makes use of the qgg and gact libraries to process and analyze GWAS (Genome-Wide Association Studies) summary statistics.

```{r all-code, include=TRUE, eval=FALSE}
# Load libraries
library(qgg)
library(gact)

# Load GAlist
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

# Check studies in gact database
GAlist$studies

# Get ldscores
ldscores <- getLDscoresDB(GAlist=GAlist, ancestry="EUR")

# Select study
studyIDs <- paste0("GWAS",c(1:2, 4:6))

# Get GWAS summary statistics from gact database
z <- getMarkerStatDB(GAlist=GAlist, studyID=studyIDs, what="z")
dim(z)

# Get number of samples for each GWAS study
n <- GAlist$studies[studyIDs,"neff"]
names(n) <- studyIDs

# Estimate narrow sense heritability using ldsc
fit <- ldsc(z=z, ldscores=ldscores, n=n, what="h2")
fit

# Estimate genetic correlation using using ldsc
fit <- ldsc(z=z, ldscores=ldscores, n=n, what="rg")
fit$h2
fit$rg
