---
title: "LD Score regression"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This script demonstrates how to estimate narrow-sense heritability and genetic correlation using the LD Score Regression (LDSC) method within the context of a gact database. The script makes use of the qgg and gact libraries to process and analyze GWAS summary statistics.

```{r all-code, include=TRUE, eval=FALSE}
# Load libraries
library(qgg)
library(gact)

# Load GAlist
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

# Check studies in gact database
GAlist$studies

# Select GWAS study IDs
studyIDs <- c("GWAS1","GWAS2","GWAS6")

# Get GWAS summary statistics for studyIDs (e.g. z and n) from gact database
stat <- getMarkerStat(GAlist=GAlist, studyID=studyIDs)

# Get ldscores matched to the ancestry of GWAS data
ldscores <- getLDscores(GAlist=GAlist, ancestry="EUR", version="1000G")

# Estimate narrow sense heritability using ldsc
fit <- ldsc(z=stat$z, n=stat$n, ldscores=ldscores, what="h2")
fit

# Estimate genetic correlation using using ldsc
fit <- ldsc(z=stat$z, n=stat$n, ldscores=ldscores, what="rg")
fit$h2
fit$rg

# Partioning of heritability using chromosome information using standard regression method
sets <- getMarkerSets(GAlist=GAlist, feature="Chromosomes")
fit <- ldsc(z=stat$z, n=stat$n, ldscores=ldscores, sets=sets, what="h2", residual=TRUE)
fit

# Partioning of heritability using regulatory categories information and a BLR method
sets <- getMarkerSets(GAlist=GAlist, feature="Regulatory Categories")
fit <- ldsc(z=stat$z, n=stat$n, ldscores=ldscores, sets=sets, what="h2", method="bayesC", residual=TRUE)
fit
