---
title: "Pathway prioritization using a BLR-MAGMA model and gene-level statistics from VEGAS"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This script is designed for advanced gene set analysis, utilizing a variety of tools and methodologies. Its primary focus lies in the integration and assessment of gene-marker sets, along with a detailed exploration of gene-disease associations derived from text mining, knowledge bases, and experimental data. The analysis utilizes gene-marker relationships, gene-level statistics, and gene-disease associations to conduct comprehensive gene set evaluations. This allow an in-depth analysis helping interpreting and validating genomic associations.

```{r all-code, include=TRUE, eval=FALSE}
# Load libraries
library(qgg)
library(gact)
library(msigdbr)
library(data.table)


# Load GAlist with information on gact database
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")


# Select studyIDs
studyIDs <- c("GWAS2","GWAS4", "GWAS5", "GWAS6")

# Study information
GAlist$studies[studyIDs,]


# Get VEGAS gene-level statistics
Z <- readRDS(file=file.path(GAlist$dirs["gsea"],"Z_vegas.rds"))
P <- readRDS(file=file.path(GAlist$dirs["gsea"],"P_vegas.rds"))

# Remove genes/rows with NA's
Z <- na.omit(Z[,studyIDs, drop=FALSE])
P <- na.omit(P[,studyIDs, drop=FALSE])

# Get gene sets for feature from mSigDB (KEGG pathways)
msigdb <- msigdbr(species = "human", category = "C2", subcategory = "CP:KEGG")
sets <- split(msigdb$ensembl_gene, f=msigdb$gs_name)

#-------------------------------------------------------------------------------
# MAGMA based on linear model (default MAGMA method)
#-------------------------------------------------------------------------------
studyIDs <- "GWAS6"
stat <- Z[, studyIDs, drop=FALSE]
fit <- qgg:::magma(stat=stat, sets=sets, method="magma", type="joint")
head(fit)

#-------------------------------------------------------------------------------
# MAGMA based on blr model (single trait)
#-------------------------------------------------------------------------------
studyIDs <- "GWAS6"
stat <- Z[, studyIDs, drop=FALSE]
fit <- qgg:::magma(stat=stat, sets=sets, method="bayesC",  pi=0.01, nit=1000, nburn=200)
head(fit)

#-------------------------------------------------------------------------------
# MAGMA based on blr model (multiple trait for males and females T2D)
#-------------------------------------------------------------------------------
studyIDs <- c("GWAS4","GWAS5")
stat <- Z[, studyIDs, drop=FALSE]
fit <- qgg:::magma(stat=stat, sets=sets, method="bayesC", pi=0.01, nit=1000, nburn=200)
str(fit)
head(fit$b)
head(fit$PIP)

#-------------------------------------------------------------------------------
# MAGMA based on blr model (multiple trait for EUR, EAS, SAS T2D)
#-------------------------------------------------------------------------------
studyIDs <- c("GWAS6","GWAS7","GWAS8")
stat <- Z[, studyIDs, drop=FALSE]
fit <- qgg:::magma(stat=stat, sets=sets, method="bayesC", pi=0.01, nit=1000, nburn=200)
str(fit)
head(fit$b)
head(fit$PIP)


#-------------------------------------------------------------------------------
# MAGMA based on blr model (multiple trait for a range of T2D related traits)
#-------------------------------------------------------------------------------
studyIDs <- c("GWAS2", "GWAS6", "GWAS9", "GWAS12","GWAS13", "GWAS14", "GWAS16",  "GWAS17", "GWAS24", "GWAS25")
stat <- Z[ ,studyIDs, drop=FALSE]
fit <- qgg:::magma(stat=stat, sets=sets, method="bayesC",  pi=0.01, nit=1000, nburn=200)
str(fit)
head(fit$b)
head(fit$PIP)


################################################################################
# Permutation test to evaluate if observed association are random
################################################################################

# Permutation to evaluate if observed PIP are non-random
stat <- Z[,"GWAS6", drop=FALSE]
stat <- apply(stat,2, function(x){sample(x)})
rownames(stat) <- rownames(Z)
fit <- qgg:::magma(stat=stat, sets=sets, method="bayesC",  pi=0.01, nit=1000, nburn=200)
head(fit)


################################################################################
# Compare to other pathway prioritization methods
################################################################################

studyID <- "GWAS6"

# GSEA based on hypergeometric test
resTHG <- gsea(stat=P[,studyID],sets=sets,method = "hyperg", threshold = 1e-8)
head(resTHG[order(resTHG$p),])

# GSEA using sum of gene test statistics
resTSUM <- gsea(stat=Z[,studyID]^2,sets=sets, nperm=100000)
head(resTSUM[order(resTSUM$p),])

# GSEA using MAGMA LM
resMAGMA_MARG <- qgg:::magma(stat=Z[,studyID], sets=sets, method="magma", type="marginal")
resMAGMA_JOINT <- qgg:::magma(stat=Z[,studyID], sets=sets, method="magma", type="joint")
rownames(resMAGMA_MARG) <- resMAGMA_MARG$ID
rownames(resMAGMA_JOINT) <- resMAGMA_JOINT$ID
head(resMAGMA_MARG)
head(resMAGMA_JOINT)

# GSEA using MAGMA BLR
resBLR <- qgg:::magma(stat=Z[,studyID], sets=sets, method="bayesC")

# Collect p-values
pvals <- data.frame(pTHG=resTHG[resBLR$ID,"p"],
              pTSUM=resTSUM[resBLR$ID,"p"],
              pMARG=resMAGMA_MARG[resBLR$ID,"p"],
              pJOINT=resMAGMA_JOINT[resBLR$ID,"p"])
rownames(pvals) <- resBLR$ID
df <- data.frame(resBLR,pvals)
head(df)

pairs(-log10(pvals))
cor(-log10(pvals))

library(UpSetR)
threshold <- 0.05
df <- as.data.frame(apply(pvals<threshold,2,as.numeric))
rownames(df) <- rownames(pvals)
head(df)
upset(data=df, text.scale=2, mainbar.y.label = "Number of associations",
      sets.x.label = "N total", sets=colnames(df),keep.order=T)

df[rowSums(df)==ncol(df),]

################################################################################ 
# GSEA based on gene-diseases association in disease database for each KEGG pathway
################################################################################

# Get gene sets for feature from mSigDB (KEGG pathways)
msigdb <- msigdbr(species = "human", category = "C2", subcategory = "CP:KEGG")
sets <- split(msigdb$ensembl_gene, f=msigdb$gs_name)

# Enrichment test based on integrated information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "DiseaseGenes", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 50)

# Enrichment test based on textmining information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "DiseaseGenesTMplus", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 50)

# Enrichment test based on textmining information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "DiseaseGenesKBplus", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 50)

# Enrichment test based on experiment information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "DiseaseGenesEXPplus", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 50)


################################################################################
# GSEA based on gene-diseases association in GWAS catalog for each KEGG pathway
################################################################################


# Enrichment test based on integrated information
res <- hgtDB(GAlist = GAlist, sets = sets, feature = "GWAScatalog", output="p") 
rws <- c("KEGG_OLFACTORY_TRANSDUCTION", 
         "KEGG_TYPE_I_DIABETES_MELLITUS", 
         "KEGG_TYPE_II_DIABETES_MELLITUS",
         "KEGG_MATURITY_ONSET_DIABETES_OF_THE_YOUNG",
         "KEGG_SYSTEMIC_LUPUS_ERYTHEMATOSUS")
cls <- grep("diabetes", tolower(colnames(res)))
head(t(res[rws,cls]), 10)

