---
title: "Polygenic scoring (PGS) using summary statistics from PGS catalog and biological pathway information"
author: "Peter Sørensen"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
The Polygenic Score (PGS) Catalog is an open database that curates polygenic scores (PGS) for various traits and diseases ([PGS Catalog](https://www.pgscatalog.org)). It provides standardized information on PGS development, validation, and performance, allowing researchers to access and compare scores across studies.

In this tutorial we show how to download marker summary statistics from PGS Catalog, compute polygenic score, and combine it with data from gact database.

<br>

#### Compute polygenic scores based on summary statistics from PGC catalog
```{r, include=TRUE, eval=FALSE}
# Load libraries
library(qgg)
library(gact)

# Load GAlist
GAlist <- readRDS(file="C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")

# Load Glist with information on 1000G matched to the ancestry of GWAS data
Glist <- readRDS(file.path(GAlist$dirs["marker"],"Glist_1000G_eas_filtered.rds"))

# Get marker summary statistics directly from PGS
stat <- fread("https://ftp.ebi.ac.uk/pub/databases/spot/pgs/scores/PGS003725/ScoringFiles/PGS003725.txt.gz",data.table=F, skip=14)

# Check column names, add marker column, and rename columns in stat object
stat$marker <- paste(stat[,"chr_name"],stat[,"chr_position"],stat[,"effect_allele"],stat[,"other_allele"],sep="_")
stat <- stat[, c("marker", "chr_name", "chr_position", "effect_allele", "other_allele", "effect_weight")]
colnames(stat) <- c("marker", "chr", "pos", "ea", "nea", "b")

# Check and align summary statistics based on marker information in Glist
stat <- checkStat(Glist=Glist, stat=stat)

# Compute polygenic scores
pgs <- gscore(Glist = Glist, stat = stat)
head(pgs)

# Extract marker sets derived from KEGG pathways
sets <- getMarkerSets(GAlist = GAlist, feature = "KEGG")

# Compute polygenic scores for KEGG pathways
pgs <- gscore(Glist=Glist, stat=stat, sets=sets)
head(pgs)
```
