---
title: "Download and prepare 1000G"
author: "Peter Sørensen"
date: "8 October 2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

Use **gbase** to prepare a Glist that selects database markers through integer
positions in the original BED files. Keep the original BED/BIM/FAM files.

```{r download}
library(gbase)
library(gact)

GAlist <- readRDS("C:/Users/gact/hsa.0.0.1/GAlist_hsa.0.0.1.rds")
GAlist <- downloadDB(GAlist=GAlist, what="1000G")
```

Prepare and save one descriptor per ancestry. This applies the website's
database-availability selection, without adding another QC filter.

```{r prepare}
for (ancestry in c("eur", "eas", "sas")) {
  prefix <- file.path(GAlist$dirs["marker"], paste0("g1000_", ancestry))
  Glist <- gprep(study=paste("1000G", toupper(ancestry)),
    bedfiles=paste0(prefix, ".bed"), bimfiles=paste0(prefix, ".bim"),
    famfiles=paste0(prefix, ".fam"))
  rsids <- intersect(unlist(Glist$rsids, use.names=FALSE), GAlist$rsids)
  Glist <- gprep(study=Glist$study, bedfiles=Glist$bedfiles,
    bimfiles=Glist$bimfiles, famfiles=Glist$famfiles, rsids=rsids)
  saveRDS(Glist, file.path(GAlist$dirs["marker"],
    paste0("Glist_1000G_", ancestry, "_filtered.rds")))
}
```

Continue with [sparse LD and LD scores](Compute_sparseLD_1000G.html).
Older qgg descriptors should be prepared again with gbase using their existing
BED/BIM/FAM paths; genotype data need not be downloaded again.
